Distance Sampling in \proglang{R}

### 2016-03-30 paper.Rmd
Jeff comments

---
author:
  - name: David L Miller
    affiliation: Integrated Statistics and University of St Andrews
    address: >
      16 Sumner Street, Woods Hole, MA 02543, USA and
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: dave@ninepointeightone.net
    url: http://converged.yt
  - name: Eric Rexstad
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: Eric.Rexstad@st-andrews.ac.uk
  - name: Len Thomas
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: len.thomas@st-andrews.ac.uk
    url: http://lenthomas.org
  - name: Laura Marshall
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: lhm@st-andrews.ac.uk
  - name: Jeffrey L Laake
    affiliation: National Oceanic and Atmospheric Administration
    address: >
      National Marine Mammal Laboratory
      Alaska Fisheries Science Center
      7600 Sand Point Way N.E., Seattle, WA 98115, USA
    email: Jeff.Laake@noaa.gov
title:
  formatted: "Distance Sampling in \\proglang{R}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Distance Sampling in R"
  # For running headers, if needed
  #short:     "\\pkg{foo}: A Capitalized Title"
abstract: >

@@

output: rticles::jss_article
bibliography: jstatsoft.bib
---

@@

# generate some animals to sample
set.seed(15) # same results every time
library("mgcv") # for inSide

@@

The columns \code{distance}, \code{Area} and (in the case of line transects) \code{Effort} have associated units (though these are not explicitly included in a \pkg{Distance} analysis). We recommended that data in these columns are converted to SI units before starting any analysis to ensure that resulting abundance and density estimates have sensible units. For example, if distances from a line transect survey are recorded in metres, the \code{Effort} columns should contain line lengths also in metres and the \code{Area} column gives the stratum areas in square metres. This would lead to density estimates of animals per square metre.

@@

In this case we may be skeptical about the top model as selected by AIC being truly better than the second best model, as there is only a very small difference in AICs. Generally, if the difference between AICs is less than 2, we may investigate multiple "best" models, potentially resorting to the simplest of these models. In the authors' experience, it is often the case that models with similar AICs will have similar estimates probabilities of detection, so in practice there is little difference in selecting between these models.

```{r summary-table, results="asis", echo=FALSE}
library("knitr")
kable(summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas),
      digits = 3, format = "latex", row.names = FALSE, escape=FALSE,
      booktabs=TRUE,
      caption="Summary for the detection function models fitted to the amakihi data. ``C-vM'' stands for Cramér-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error. Models are sorted according to AIC.\\label{tab:amakihi}")
```

@@

We combine these two variances by noting that squared coefficients of variation (approximately) add [@Goodman:1960dd] (often referred to as "the delta method").

@@

The examples of applications for which distance sampling has been used are many.  Objects being detected need not be animals; plants [@Buckland:2007kl], beer cans [@Otto:1990gs], elephant dung [@NchPlu01] and bricks on a lake bottom [@Ber90] have been subjects of distance sampling investigations. The method of detection need not be visual sightings. Detections of objects can be through auditory as well as visual means [@Marques2013b]. Songs of birds [@Buckland2006] and whale vocalisations [@BorSub1] are just two examples. Blows made by whales are examples of processes that can be modelled using distance sampling. Songs and blows are indirect sampling methods producing estimates of cue density. Cue densities can be converted to animal densities with additional data needed to estimate the rate at which cues are produced and the rate at which they disappear. These are but a few examples of the applications to which distance sampling has been applied (a larger, but incomplete list of references is given at \url{http://distancesampling.org/dbib.html}).

@@

In combination with tools such as \pkg{knitr} and \pkg{rmarkdown} [@rmarkdown-pkg], the helper functions in \pkg{Distance} provide a useful set of tools to perform reproducible analyses of wildlife abundance for both managers and ecologists. \proglang{R} and its extension packages provide many tools exploratory data analysis that can be useful for a distance sampling analysis. We hope that this paper provides useful examples for those wishing to pursue distance sampling in \proglang{R}. More information on distance sampling can be found at \url{http://distancesampling.org} and a mailing list is maintained at \url{https://groups.google.com/forum/#!forum/distance-sampling}.

>>>>

---
author:
  - name: David L Miller
    affiliation: Integrated Statistics and University of St Andrews
    address: >
      16 Sumner Street, Woods Hole, MA 02543, USA and
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: dave@ninepointeightone.net
    url: http://converged.yt
  - name: Eric Rexstad
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: Eric.Rexstad@st-andrews.ac.uk
  - name: Len Thomas
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: len.thomas@st-andrews.ac.uk
    url: http://lenthomas.org
  - name: Laura Marshall
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: lhm@st-andrews.ac.uk
  - name: Jeffrey L Laake
    affiliation: National Oceanic and Atmospheric Administration
    address: >
      Marine Mammal Laboratory
      Alaska Fisheries Science Center
      7600 Sand Point Way N.E., Seattle, WA 98115, USA
    email: Jeff.Laake@noaa.gov
title:
  formatted: "Distance Sampling in \\proglang{R}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Distance Sampling in R"
  # For running headers, if needed
  #short:     "\\pkg{foo}: A Capitalized Title"
abstract: >

@@

output:
  rticles::jss_article:
    citation_package: natbib
bibliography: jstatsoft.bib
---

@@

# generate some animals to sample
set.seed(15) # same results every time
library("mgcv", quietly=TRUE) # for inSide

@@

The columns \code{distance}, \code{Area} and (in the case of line transects) \code{Effort} have associated units (though these are not explicitly included in a \pkg{Distance} analysis). We recommend that data in these columns are converted to SI units before starting any analysis to ensure that resulting abundance and density estimates have sensible units. For example, if distances from a line transect survey are recorded in metres, the \code{Effort} columns should contain line lengths also in metres and the \code{Area} column gives the stratum areas in square metres. This would lead to density estimates of animals per square metre.

@@

In this case we may be skeptical about the top model as selected by AIC being truly better than the second best model, as there is only a very small difference in AICs. Generally, if the difference between AICs is less than 2, we may investigate multiple "best" models, potentially resorting to the simplest of these models. In the authors' experience, it is often the case that models with similar AICs will have similar estimates probabilities of detection, so in practice there is little difference in selecting between these models. It is important to note that comparing AICs between models with different truncations is not appropriate, as models with different truncation use different data.

```{r summary-table, results="asis", echo=FALSE}
kable(summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas),
      digits = 3, format = "latex", row.names = FALSE, escape=FALSE,
      booktabs=TRUE,
      caption="Summary for the detection function models fitted to the amakihi data. ``C-vM'' stands for Cramér-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error. Models are sorted according to AIC.\\label{tab:amakihi}")
```

@@

We combine these two variances by noting that squared coefficients of variation (approximately) add [often referred to as "the delta method"; @seber2002estimation].

@@

Distance sampling has been applied in a wide variety of situations. Objects being detected need not be animals; plants [@Buckland:2007kl], beer cans [@Otto:1990gs], elephant dung [@NchPlu01] and bricks on a lake bottom [@Ber90] have been subjects of distance sampling investigations. The method of detection need not be visual sightings. Detections of objects can be through auditory as well as visual means [@Marques2013b]. Songs of birds [@Buckland2006] and whale vocalisations [@BorSub1] are just two examples. Blows made by whales are examples of processes that can be modelled using distance sampling. Songs and blows are indirect sampling methods producing estimates of cue density. Cue densities can be converted to animal densities with additional data needed to estimate the rate at which cues are produced and the rate at which they disappear. These are but a few examples of the applications to which distance sampling has been applied (a larger, but incomplete list of references is given at \url{http://distancesampling.org/dbib.html}).

@@

In combination with tools such as \pkg{knitr} and \pkg{rmarkdown} [@rmarkdown-pkg], the helper functions in \pkg{Distance} provide a useful set of tools to perform reproducible analyses of wildlife abundance for both managers and ecologists. We also direct readers' attention to \pkg{DsShiny} [@dsshiny-pkg], a package that builds on \pkg{Shiny} [@shiny-pkg] and \pkg{mrds} to allow users to build and fit models in a graphical interface, with output to RMarkdown.

\proglang{R} and its extension packages provide many tools exploratory data analysis that can be useful for a distance sampling analysis. We hope that this paper provides useful examples for those wishing to pursue distance sampling in \proglang{R}. More information on distance sampling can be found at \url{http://distancesampling.org} and a mailing list is maintained at \url{https://groups.google.com/forum/#!forum/distance-sampling}.
